---
title: "survival.analysis"
author: "Team 4"
date: "2026-01-23"
output: html_document
---

```{r}
library(survival)
library(foreign)
library(ggplot2)
library(survminer)
library(rms)
library(flexsurv)
library(dplyr)
library(ciTools)
library(here)
library(visreg)
library(cmprsk)
library(reticulate)
library(MASS)
library(tidyr)

use_python("C:\\ProgramData\\Anaconda3\\python.exe")
```


```{r}
library(aws.s3)

#Finding the AWS S3 bucket
bucket_exists(
  bucket = "s3://survival2024/",
  region = "us-east-1"
)
```

```{r}
files <- get_bucket_df(
  bucket = "s3://survival2024/",
  region = "us-east-1",
  max = 20000
) %>%
  as_tibble()

save_object(
  object = "hurricane.csv", 
  bucket = "s3://survival2024/", 
  region = "us-east-1", 
  file = "hurricane_file"
)

# Read it into a new dataframe called 'pumps'
pumps <- read.csv("hurricane_file")

```
```{r}
colnames(pumps)
```




For this phase use the entire data set.
Build a Cox proportional hazard model for motor failure with main effects only using
backward elimination.
```{r}
#create event motor failure variable
pumps <- pumps |>
  mutate(motor_event = ifelse(reason == 2, 1, 0))

#run backward selection
full.model <- coxph(Surv(hour, motor_event) ~ backup + age + bridgecrane + 
                       servo + gear + trashrack + slope + elevation, data=pumps)

n <- nrow(pumps)
bic.model <- stepAIC(full.model, direction="backward", k=log(n))

summary(bic.model)

```

```{r}
#confirm quasi complete separation between trashrack and motor_event
table(pumps$trashrack, pumps$motor_event)

#remove trashrack from backward selection model
#run backward selection
final.model <- coxph(Surv(hour, motor_event) ~ age + slope, data=pumps)
summary(final.model)
```

The Army Corp of Engineers and the Committee are interested in finding out if this type of
failure is more likely to occur if the motor is constantly running for 12 hours blocks.
```{r}
#adjust to counting long format (one row per hour)
pumps_long <- pumps |>
  mutate(id = row_number()) |>
  dplyr::select(id, hour, motor_event, age, trashrack, slope, h1:h48) |> #drop vars not in model
  pivot_longer(
               cols = c(starts_with("h") , -hour),
               names_to = "time_point",
               values_to = "is_running") |>
  mutate(time_point = as.numeric(gsub("h", "", time_point))) |>
  filter(time_point <= hour) |> #remove rows after failure
  arrange(id, time_point)

#create flag for 12-hr block
pumps_long <- pumps_long |>
  group_by(id) |>
  mutate(
    run_streak = sequence(rle(is_running)$lengths) * is_running, #if is_running is 0, streak resets, else it adds up
    twelve.hr = ifelse(run_streak >=12, 1, 0), #accounts for 12+ hrs consecutive run
    tstart = time_point - 1, #beginning of the single hour
    tstop = time_point, #end of the single hour
    event_status = ifelse(time_point == hour & motor_event == 1, 1, 0)
        ) |>
  ungroup()


#build time-dependent Cox PH model
time.model <- coxph(Surv(tstart, tstop, event_status) ~ age + slope + twelve.hr, 
                  data = pumps_long)


summary(time.model) #We find that twelve.hr flag is NOT significant

```


Test assumptions for model
```{r}
#Schoenfeld residuals to test for proportional hazards
ph_test <- cox.zph(time.model)
print(ph_test)
```
```{r}
library(survminer)

# ggcoxfunctional automatically checks the linearity of continuous covariates
# It fits a null model and plots the variable against the martingale residuals
ggcoxfunctional(final.model, data = pumps)

log.model <- coxph(Surv(hour, motor_event) ~ log(age) + slope, data=pumps)
summary(log.model)
ggcoxfunctional(log.model, data = pumps)
```

```{r}
ggcoxfunctional(time.model, data = pumps_long)
```

Bin continuous variables that don't meet linearity assumption
```{r}
hist(pumps$age)
max(pumps$age)

#create the quartiles for age and slope
pumps$age_bin <- cut(pumps$age, 
                     breaks = quantile(pumps$age, probs = seq(0, 1, 0.25), na.rm = TRUE),
                     include.lowest = TRUE, 
                     labels = c("Q1_young", "Q2", "Q3", "Q4_old"))
pumps$slope_bin <- cut(pumps$slope, 
                     breaks = quantile(pumps$slope, probs = seq(0, 1, 0.25), na.rm = TRUE),
                     include.lowest = TRUE, 
                     labels = c("Q1_young", "Q2", "Q3", "Q4_old"))

#run cox model with the binned variable
bin.model <- coxph(Surv(hour, motor_event) ~ age_bin + slope_bin, data = pumps)

# 3. View the Hazard Ratios
summary(bin.model)

```
Test proportional hazards for binned model
```{r}
#Schoenfeld residuals to test for proportional hazards
ph_test <- cox.zph(bin.model)
print(ph_test)

#high p-value -> fail to reject -> hazards ARE proportional
```
Re-do time dependent model
```{r}
#adjust to counting long format (one row per hour)
pumps_long <- pumps |>
  mutate(id = row_number()) |>
  dplyr::select(id, hour, motor_event, age_bin, trashrack, slope_bin, h1:h48) |> #drop vars not in model
  pivot_longer(
               cols = c(starts_with("h") , -hour),
               names_to = "time_point",
               values_to = "is_running") |>
  mutate(time_point = as.numeric(gsub("h", "", time_point))) |>
  filter(time_point <= hour) |> #remove rows after failure
  arrange(id, time_point)

#create flag for 12-hr block
pumps_long <- pumps_long |>
  group_by(id) |>
  mutate(
    run_streak = sequence(rle(is_running)$lengths) * is_running, #if is_running is 0, streak resets, else it adds up
    twelve.hr = ifelse(run_streak >=12, 1, 0), #accounts for 12+ hrs consecutive run
    tstart = time_point - 1, #beginning of the single hour
    tstop = time_point, #end of the single hour
    event_status = ifelse(time_point == hour & motor_event == 1, 1, 0)
        ) |>
  ungroup()

#build time-dependent Cox PH model with binned variables
time.model <- coxph(Surv(tstart, tstop, event_status) ~ age_bin + slope_bin + twelve.hr, 
                  data = pumps_long)


summary(time.model) #We find that twelve.hr flag is NOT significant
```













