---
title: "Fraud HW 1"
author: "Gray Nico"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readr)
library(MASS)
library(dbscan)

df_raw <- read_csv("aggregated_policy_nolabel_2026.csv")

# Necessary variable cleaning
df <- df_raw %>%
  mutate(
    Birthday = as.Date(Birthday, format = "%d-%b-%y"),
    Birthday = if_else(Birthday > Sys.Date(), Birthday - lubridate::years(100), Birthday),
    ZipCode = as.character(ZipCode),
    Diff_Cust_Inc_CZIPInc = as.numeric(Diff_Cust_Inc_CZIPInc),
    Ratio_CL_PopZIP = as.numeric(Ratio_CL_PopZIP),
    Ratio_AdjZIP_PopZIP = as.numeric(Ratio_AdjZIP_PopZIP),
    Med_Conditions = as.numeric(Med_Conditions)
  ) %>%
  
  # FEETINCHES CLEANING
  mutate(FeetInches = gsub("['\"]", "", FeetInches)) %>%  # strip ' and "
  separate(FeetInches, into = c("feet", "inches"), sep = "\\s+", convert = TRUE) %>%
  mutate(feet_inches = feet + inches / 10) %>%
  dplyr::select(-feet, -inches)

```

## Data Exploration

The code below creates 6 flags that can be used to pull out IDs that have suspicious
values in multiple fields. IDs that were flagged have been saved in separate
vectors for comparisons.

- **Mismatched Zipcode:** the given city-zipcode match is not identical to a verified
list of city-zipcode pairings.
  - Note that this is an extremely aggressive flag, and there are many cases
  where the provided city was simply another version or abbreviation for the
  verified city
- **Missing Zipcode:** the given zipcode was not found in a verified dataset of
USA zipcodes.
  - Note that this may incorrectly flag zipcodes belonging to another country.
- **Suspicious Birthday:** provided birthday was on January 1st, or on the 1st
or 15th of another month.
  - Note that this is very aggressive; this should only be used along with many
  other flags for consideration.
- **Suspicious Telephone:** provided phone number matches at least one suspicious
pattern (e.g., 000, 999, invalid area code, 7+ repeated digits, sequential, etc.)
- **Suspicious Amount:** reward amount was an outlying value, given the policy type and reward reason
- **Suspicious Income:** customer has a very large income compared to zipcode median, with an unusually high claim rate compared to zipcode rate, and experienced a very high change in income compared to zipcode population
- **Suspicious Claim Behavior:** customer frequently changed coverage limits with a high recent volatility, and claimed a reward shortly after last change

```{r Check Zipcode Matches}

zipcodes <- read_csv("uszips.csv")

# Left join table A with table B on the zip code
df <- df %>%
  left_join(zipcodes %>% dplyr::select(zip, city), 
            by = c("ZipCode" = "zip")) %>%
  mutate(
    zip_flag = case_when(
      is.na(city)          ~ 2L,   # Zip not found in table B at all
      City != city         ~ 1L,   # Zip found but city doesn't match
      TRUE                 ~ 0L    # All good
    ),
    zip_city_matches = if_else(
      !is.na(city),
      paste0(City, " | ", city),   # "Table A city | Table B city"
      paste0(City, " | N/A")       # Zip wasn't in table B
    )
  ) %>%
  dplyr::select(-city)  # Drop the lookup column from table B

# --- Extract Cust_IDs ---

# Cust_IDs where city didn't match (zip exists in B but wrong city)
mismatch_ids <- df %>%
  filter(zip_flag == 1) %>%
  pull(Cust_ID)

# Cust_IDs where zip wasn't found in table B at all
missing_zip_ids <- df %>%
  filter(zip_flag == 2) %>%
  pull(Cust_ID)

# View rows and relevant columns where the flag was 1
df %>%
  filter(zip_flag %in% c(1)) %>%
  dplyr::select(Cust_ID, zip_city_matches)

```

```{r Check Birthdays}

# --- Check 1: Day-of-month distribution (rounding/fabrication signal) ---
# Fabricated birthdays cluster heavily on the 1st, 15th, round numbers
df %>%
  mutate(day = as.integer(format(as.Date(Birthday), "%d"))) %>%
  count(day) %>%
  ggplot(aes(x = day, y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Births by Day of Month", x = "Day", y = "Count") +
  theme_minimal()

# --- Check 2: Month distribution (should be roughly uniform-ish) ---
df %>%
  mutate(month = as.integer(format(as.Date(Birthday), "%m"))) %>%
  count(month) %>%
  ggplot(aes(x = factor(month), y = n)) +
  geom_col(fill = "steelblue") +
  labs(title = "Births by Month", x = "Month", y = "Count") +
  theme_minimal()

# --- Check 3: Year distribution (spot WWII dip, Baby Boom spike) ---
df %>%
  mutate(year = as.integer(format(as.Date(Birthday), "%Y"))) %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_line(color = "steelblue") +
  labs(title = "Births by Year", x = "Year", y = "Count") +
  theme_minimal()

# --- Check 4: Flag suspicious birthdays (1st of month, Jan 1st, etc.) ---
df <- df %>%
  mutate(
    bday_day   = as.integer(format(as.Date(Birthday), "%d")),
    bday_month = as.integer(format(as.Date(Birthday), "%m")),
    bday_flag  = case_when(
      bday_month == 1 & bday_day == 1 ~ "Jan 1st (likely placeholder)",
      bday_day == 1                   ~ "1st of month (possible rounding)",
      bday_day == 15                  ~ "15th of month (possible rounding)",
      TRUE                            ~ "OK"
    )
  )

susp_bday_ids <- df %>%
  filter(bday_flag != "OK") %>%
  pull(Cust_ID)

```

```{r Check Telephone Numbers}

df <- df %>%
  mutate(
    phone_clean = gsub("[^0-9]", "", TelephoneNumber),
    phone_flag = case_when(
      nchar(phone_clean) != 10                                                          ~ 1L,  # wrong length
      grepl("^555", phone_clean)                                                        ~ 1L,  # 555 numbers
      grepl("^(000|911|999)", phone_clean)                                              ~ 1L,  # invalid area codes
      grepl("(\\d)\\1{6,}", phone_clean)                                               ~ 1L,  # 7+ repeated digits
      grepl("^(1234567890|0987654321)", phone_clean)                                   ~ 1L,  # sequential
      substr(phone_clean, 1, 3) %in% c("000", "100", "200", "300", "400", "500") ~ 1L,  # unlikely area codes
      substr(phone_clean, 4, 6) %in% c("555", "000", "111")                           ~ 1L,  # suspicious exchanges
      TRUE                                                                              ~ 0L
    )
  )

susp_phone_ids <- df %>%
  filter(phone_flag == 1) %>%
  pull(Cust_ID)

```

```{r Checking Physical Implausibility - UNUSED}

df <- df %>%
  mutate(
    feet            = floor(feet_inches),
    inches          = round((feet_inches - floor(feet_inches)) * 10),
    height_in_total = feet * 12 + inches,
    bmi             = (Pounds / height_in_total^2) * 703,
    physical_flag   = case_when(
      Pounds < 90 | Pounds > 400                   ~ 1L,  # outside broad healthy adult range
      height_in_total < 54 | height_in_total > 84  ~ 1L,  # outside 4'6" to 7'0"
      bmi < 13 | bmi > 55                          ~ 1L,  # outside survivable BMI range
      TRUE                                         ~ 0L
    )
  ) %>%
  dplyr::select(-feet, -inches, -height_in_total, -bmi)

susp_bmi_ids <- df %>%
  filter(physical_flag == 1) %>%
  pull(Cust_ID)

# No suspicious BMIs

```

```{r Checking Reward Amount Anomalies}

df <- df %>%
  group_by(Type, Reward_Reason) %>%
  mutate(
    reward_mean = mean(Reward_Amount, na.rm = TRUE),
    reward_sd   = sd(Reward_Amount, na.rm = TRUE),
    reward_z    = (Reward_Amount - reward_mean) / reward_sd,
    reward_flag = if_else(abs(reward_z) > 3, 1L, 0L)
  ) %>%
  ungroup() %>%
  dplyr::select(-reward_mean, -reward_sd, -reward_z)

susp_amt_ids <- df %>%
  filter(reward_flag == 1) %>%
  pull(Cust_ID)

```

```{r Checking Income Inconsistencies}

df <- df %>%
  mutate(
    geo_income_flag = case_when(
      abs(Diff_Cust_Inc_CZIPInc) > 2 * sd(Diff_Cust_Inc_CZIPInc, na.rm = TRUE) &
        Ratio_CL_PopZIP > quantile(Ratio_CL_PopZIP, 0.95, na.rm = TRUE)  ~ 1L,  # high claim ZIP + income outlier
      abs(Zscore_Inc_Change) > 3                                                    ~ 1L,  # extreme income change
      TRUE                                                                      ~ 0L
    )
  )

susp_income_ids <- df %>%
  filter(geo_income_flag == 1) %>%
  pull(Cust_ID)

```

```{r Checking Coverage Changes}

df <- df %>%
  mutate(
    churn_flag = case_when(
      Number_Changes > quantile(Number_Changes, 0.95, na.rm = TRUE) &
        SD_CovLim_Last_6mon > quantile(SD_CovLim_Last_6mon, 0.95, na.rm = TRUE) &
        Time_Between_CH_R < 90                                            ~ 1L,  # churned + claimed fast
      Time_Between_CH_R < 30 & Number_Changes >= 3                   ~ 1L,  # very fast claim after repeated changes
      TRUE                                                                     ~ 0L
    )
  )

susp_claims_ids <- df %>%
  filter(churn_flag == 1) %>%
  pull(Cust_ID)

```

```{r Scoring IDs}

# --- Step 1: Tag each ID with its source vector ---
reasonable_flags <- bind_rows(
  data.frame(Cust_ID = mismatch_ids,    flag = "mismatch_ids"),
  data.frame(Cust_ID = missing_zip_ids, flag = "missing_zip_ids"),
  data.frame(Cust_ID = susp_bday_ids,   flag = "susp_bday_ids"),
  data.frame(Cust_ID = susp_phone_ids,  flag = "susp_phone_ids")
)

concerning_flags <- bind_rows(
  data.frame(Cust_ID = susp_amt_ids,    flag = "susp_amt_ids"),
  data.frame(Cust_ID = susp_income_ids, flag = "susp_income_ids"),
  data.frame(Cust_ID = susp_claims_ids, flag = "susp_claims_ids")
)

# --- Step 2: Count flags per ID by category ---
reasonable_counts <- reasonable_flags %>%
  distinct(Cust_ID, flag) %>%          # deduplicate in case an ID appears twice in same vector
  group_by(Cust_ID) %>%
  summarise(
    n_reasonable     = n(),
    reasonable_flags = paste(flag, collapse = ", "),
    .groups = "drop"
  )

concerning_counts <- concerning_flags %>%
  distinct(Cust_ID, flag) %>%
  group_by(Cust_ID) %>%
  summarise(
    n_concerning     = n(),
    concerning_flags = paste(flag, collapse = ", "),
    .groups = "drop"
  )

# --- Step 3: Merge and fill zeros for IDs missing from either category ---
flag_summary <- full_join(reasonable_counts, concerning_counts, by = "Cust_ID") %>%
  mutate(
    n_reasonable     = replace_na(n_reasonable, 0),
    n_concerning     = replace_na(n_concerning, 0),
    reasonable_flags = replace_na(reasonable_flags, "none"),
    concerning_flags = replace_na(concerning_flags, "none")
  )

# --- Step 4: Score ---
flag_summary <- flag_summary %>%
  mutate(
    fraud_score = case_when(
      n_concerning == 3 & n_reasonable >= 2  ~ 10L,
      n_concerning == 3 & n_reasonable == 1  ~  9L,
      n_concerning == 3 & n_reasonable == 0  ~  8L,
      n_concerning == 2 & n_reasonable >= 2  ~  7L,
      n_concerning == 2 & n_reasonable == 1  ~  6L,
      n_concerning == 2 & n_reasonable == 0  ~  5L,
      n_concerning == 1 & n_reasonable >= 3  ~  4L,
      n_concerning == 1 & n_reasonable %in% c(1, 2) ~  3L,
      n_concerning == 1 & n_reasonable == 0  ~  2L,
      n_concerning == 0 & n_reasonable >= 2  ~  1L,
      n_concerning == 0 & n_reasonable == 1  ~  0L,
      TRUE                                   ~  NA_integer_  # only 1 reasonable flag; no action needed
    )
  ) %>%
  arrange(desc(fraud_score))

# IDs with a score of 5 or higher
high_susp_ids <- flag_summary %>% 
  filter(fraud_score >= 5) %>%
  pull(Cust_ID)

```

## LOF Analysis

This section uses the code provided by Dr. Simmons to use the Local Outlier Factor (LOF) on multiple variables. 

```{r Data Prepping}

# Ensuring that all numeric variables are handled properly
cols <- data.frame(
  column = character(),
  type = character()
)
for (col in colnames(df)){
  cols <- rbind(cols, data.frame(column = col, type = class(df[[col]])[1]))
}

chars <- cols %>% filter(type == "character")
nums <- cols %>% filter(type == "numeric")

# Creating a df with numeric ccolumns of interest
df_nums <- df %>%
  dplyr::select(
    Cust_ID,
    Reward_Amount, # *
    Number_Changes, # *
    Cov_Limit_Claim, # *
    Income_Claim, # *
    Adj_Tech_Confidence, # *
    Tech_Adj_Confidence, # *
    Coverage_Income_Ratio_Claim, # *
    Coverage_Income_Ratio_Initial, # *
    Time_Between_CL_R, # *
    Time_Between_IN_CL, # * 
    Med_Conditions, # *
    Pol_Count, # *
    Ratio_CL_PopZIP,
    Zscore_Inc_Change, # *
    CH_per_Year, # *
    Time_Between_CH_R, # *
    PerSize_Last_CH, # *
    Ratio_CH_TimeCompany, # *
    Diff_Cust_Inc_CZIPInc,
    Ratio_AdjZIP_PopZIP, # *
    SD_CovLim_Last_6mon # *
  )

df_clean <- na.omit(df_nums)

```

```{r LOF}

# Separate ID from numeric columns
ids <- df_clean$Cust_ID
df_scale <- df_clean[, !colnames(df_clean) %in% "Cust_ID"]

# Remove zero-variance columns and scale
df_scale <- df_scale[, sapply(df_scale, function(x) sd(x) > 0)]
lof_scores <- lof(scale(df_scale), minPts = 5)

# Attaching scores
df_scale$lof_score <- lof_scores

# Attaching ID
df_scale$ID <- ids

# Outputting IDs with a LOF score of 1.35 or higher

susp_LOF_ids <- df_scale %>%
  filter(lof_score >= 1.35) %>%
  pull(ID)

```

This code produced a new vector of IDs based on LOF score. 1319 customers had a score of 1.35 or higher.

As a final check, the below code saves IDs identified as both highly suspicious from the custom flags, AND identified as highly suspicious from the LOF scoring. Only 66 fall into this group, but they are very likely to be fraudulent.

```{r Final Check}

final_ids <- intersect(high_susp_ids, susp_LOF_ids)

```



